Regression models for Cylindrical data in Psychology

### 2019-01-28 Manuscript/ManuscriptMarkdownCV.Rmd
Update manuscript

---
title: "Regression models for Cylindrical data in Psychology"
author: #|
 # | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
 # | $^1$Department of Methodology and Statistics, Utrecht University
 # | $^2$TNO
  #| $^3$Department of Education, Utrecht University
  #| $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}
  - \setlength\parindent{24pt}
  - \usepackage[nodisplayskipstretch]{setspace}\doublespacing
  #- \usepackage{endfloat}
  - \DeclareRobustCommand{\VANDER}[3]{#2} % set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  Cylindrical data are multivariate data which consist of a directional,
  in this paper circular, and a linear component. Examples of cylindrical
  data in psychology include human navigation (direction and distance of 
  movement), eye-tracking research (direction and length of saccades) and
  data from an interpersonal circumplex (type and strength of
  interpersonal behavior). In this paper we adapt four models for
  cylindrical data to include a regression of the circular and linear
  component onto a set of covariates. Subsequently, we illustrate how to
  fit these models and interpret their results on a dataset on the
  interpersonal behavior of teachers.
---
```{r inlude = FALSE, echo = FALSE}

@@

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
library(caret) #cross-validation
```

@@

Cylindrical data are data that consist of a linear variable and a directional
variable. In this paper, the directional variable is circular, meaning that it
consists of a single angle instead of a set of angles. A circular variable is
different from a linear variable in the sense that it is measured on a different
scale. Figure \ref{circline} shows the difference between a circular scale
(right) and a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in
fact represent the same number while on a linear scale the two ends, \(-\infty\)
and \(\infty\), are not connected and consequently the values 0\(^\circ\) and
360\(^\circ\) are located on different places on the scale. This difference
requires us to use special statistical methods for circular variables (see e.g.
@fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
  elaborate overview). For instance, a notion as simple as the sample average
  needs to be re-defined. As is the case for circular data, the analysis of
  cylindrical data also requires special methods.\newline
\indent Cylindrical data occur in several fields of research, such as for
instance in meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]. However, also in psychology several types
of cylindrical data can be found. For example, in research on human navigation
in the field of cognitive psychology both the distance, a linear variable, and
the direction, a circular variable, of movement are of interest
[@chrastil2017rotational]. In eye-tracking research, saccade data are an example
of cylindrical data, because both the direction (\emph{i.e.}, the circular variable)
and the duration (\emph{i.e.}, the linear variable) of the saccades are of interest
(for a review of eye-tracking research see  @rayner200935th).\newline
\indent The type of data that is used in the present study are also
psychological, namely data from circumplex measurement instruments. For
instance, data from the interpersonal circumplex as used in personality
psychology are by definition of a cylindrical nature (see Section \ref{Example}
for a more detailed explanation).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.\newline
\indent Our aim in this paper is to fill this gap in the literature by adapting
four existing cylindrical models in such a way that they include a regression of
both the linear and circular component of a cylindrical variable onto a set of
covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. Additionally, we will show how a
correct statistical treatment of such cylindrical data can lead to new insights.
We will do this for the teacher data, a dataset from the field of educational
psychology. In the teacher data, apart from modelling the relation between the
linear and circular component of a cylindrical variable we would also like to
predict the two components from a set of covariates in a regression model.

The paper is organized as follows: Section \ref{Example} describes the teacher
data, while Section \ref{Models} presents the four cylindrical models and our
associated adaptations to new regression models. In that same section we also
discuss the model fit criterion that we will use in Section \ref{DataAnalysis}
for the comparison of the four models. A detailed data analysis with interesting
new insights is also provided in Section \ref{DataAnalysis}. We conclude the
paper with a discussion in Section \ref{Discussion}, and the Supplementary Material collects
the technical details of the MCMC procedures.

@@

\section{Teacher data}\label{Example}

@@

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
@vanderWant2015role, @Claessens2016side and @pennings2018interpersonal. An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see @horowitz2010handbook for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. This two-dimensional space is called the interpersonal
circle/circumplex (IPC). The IPC is ``a continuous order with no beginning or
end'' [@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering
and the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular
variable.\newline
\indent @Cremers2018Assessing explain the circular nature of the IPC data and
analyze them as such using a circular regression model. The two-dimension scores
Agency and Communion can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}, where \(A\) represents a score on the
Agency dimension and \(C\) represents a score on the Communion dimension
\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
\end{equation}
The resulting circular variable \(\theta\) can then be modelled and takes values
in the interval $[0, 2\pi)$. However, when two-dimensional data are converted
to the circle we lose some information, namely the length of the two-dimensional
vector \((A, C)^t\), \emph{i.e.}, its Euclidean norm \(\mid\mid (A, C)^t \mid\mid\).
This length represents the strength of the type of interpersonal behavior a
teacher shows towards his/her students and can be considered as the linear
variable in a cylindrical model, allowing us to model a circular variable
\(\theta\) together with the linear variable corresponding to \(\mid\mid (A, C)^t
\mid\mid\). This leads to an improved analysis of interpersonal circumplex data
as we take all information into account. In the next section we introduce
several models that can be used for a more accurate and informative regression
analysis on the teacher data. First however we will provide descriptives for our
data set.

\subsection{Data description}\label{DataDescriptives}

```{r loaddata, return = FALSE, echo = FALSE}
Dat <- read_spss("Data Heleen/mergeddata.sav")

@@

@@

plot(Dat$y, Dat$theta*(180/pi), 
     ylab = "IPC location", xlab = "IPC strength",
     bty = "n", ylim = c(-180, 180))

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. In addition to the score on the IPC, the
circular outcome, and the strength of the score on the IPC, the linear outcome,
a teachers' self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. After listwise deletion of missings ($3$ in total,
only for the self-efficacy) we have a sample of 148 teachers. Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset. Here
$\hat{\rho}$ is a sample estimate for the circular concentration where a value
of 0 means that the data is not concentrated at all, \emph{i.e.} spread over the
entire circle, and a value of 1 means that all data is concentrated at a single
point on the circle. Figure \ref{dataplot} is a scatterplot showing the relation
between the linear and circular outcome of the teacher data.

@@

\begin{table}[h]
\centering
\caption{Descriptives for the teacher dataset.} 
\begin{tabular}{lrrrl}
  \noalign{\smallskip}\hline\noalign{\smallskip}
Variable & mean/$\bar{\theta}$ & sd/$\hat{\rho}$ & Range & Type \\ \hline\noalign{\smallskip}
IPC &33.22$^\circ$& 0.76 & - & Circular\\
strength IPC & 0.43 & 0.15 & 0.08 - 0.80 & Linear\\
SE & 5.04 & 1.00 & 1.5 - 7.0 & Linear\\
   \hline
\end{tabular}
\label{Tableteacherdescriptives}
\end{table}

\section{Four cylindrical regression models}\label{Models}

In this section we present four cylindrical models and adapt them such
that they contain predictors for the linear and circular outcomes, \(Y\)
and \(\Theta\). The first two models are based on a construction by
@mastrantonio2015bayesian, while the other models are extensions of the
models from @abe2017tractable and @mastrantonio2018joint.

\subsection{The modified CL-PN and modified CL-GPN  models}\label{CL-(G)PN}

Following @mastrantonio2015bayesian we consider in this section two models where
the relation between \(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\)
and $q$ covariates is specified as
\begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation}
\noindent where the random variable \(R\geq0\) will be introduced below, the
error term \(\epsilon \sim N(0, \sigma^2)\) with variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are the
intercept and regression coefficients and \(x_1, \dots, x_q\) are the \(q\)
covariates. In both of these models the
conditional distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given
by
\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation}
\noindent where \(c = \begin{bmatrix} r \cos(\theta) \\ r\sin(\theta)
\end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\). The linear outcome thus has a normal distribution conditional on
\(\Theta\) and \(R\) and contains already linear covariates \(x_1, \dots, x_q\)
in its location part. For the teacher dataset, the regression equation for the
linear outcome in the CL-PN and CL-GPN model is the following: \[\hat{y}_i =
\gamma_0 + \gamma_{cos}\cos(\theta_i)r_i + \gamma_{sin}\sin(\theta_i)r_i +
\gamma_1\text{SE}_i,\]
\noindent where $\text{SE}_i$ is the self-efficacy score of one individual $i = 1, \dots, n$ where $n$ is the sample size.\newline
\indent For the circular outcome we assume either a projected normal (PN) or a general
projected normal (GPN) distribution. These distributions arise from the radial
projection of a distribution defined on the plane onto the circle. The relation
between a bivariate vector \(\boldsymbol{S}\) in the plane and the circular
outcome \(\Theta\) is defined as follows
\begin{equation}\label{projection}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = R\boldsymbol{u} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation}
\noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\), the Euclidean norm of
the bivariate vector \(\boldsymbol{S}\). In the PN distribution we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\) and in the GPN we
assume \(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
$\boldsymbol{\mu} \in \mathbb{R}^2$, \(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1
\end{bmatrix}\), \(\rho \in (-\infty, +\infty)\) and \(\tau^2 \geq 0\) (as in
@hernandez2016general). This leads to the
circular-linear PN (CL-PN) and circular-linear GPN (CL-GPN) distributions. We
will now detail how we modify both cylindrical distributions to also incorporate
covariates for the circular part.

\subsubsection{The modified CL-PN distribution}

Following @nunez2011bayesian, the joint density of \(\Theta\) and \(R\) for the
PN distribution equals
\begin{equation}\label{pnreg}
f(\theta,r \mid \boldsymbol{\mu}, \boldsymbol{I}) = \frac{r}{2\pi} \exp\left[- \frac{(r\boldsymbol{u} - \boldsymbol{\mu})^t(r\boldsymbol{u} - \boldsymbol{\mu})}{2}\right],
\end{equation}
\noindent where \(\boldsymbol{u}= \begin{bmatrix} \cos (\theta) \\ \sin (\theta)
\end{bmatrix}\) and $r$ is the same as in \eqref{circlinlink} and
\eqref{ycondtheta} and is defined in \eqref{projection}. In a regression setup
the outcomes \(\theta_i,r_i\) for each individual \(i = 1, \dots, n\), where
\(n\) is the sample size, are generated independently from the distribution with
density \eqref{pnreg}. The mean vector \(\boldsymbol{\mu}_i \in \mathbb{R}^2\)
is then defined as \(\boldsymbol{\mu}_i = \boldsymbol{B}^t\boldsymbol{z}_i\)
where the vector \(\boldsymbol{z}_i\) is a vector of dimension $p + 1$ that
contains the covariate values and the value 1 to estimate an intercept and
\(\boldsymbol{B} = (\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})\) contains
the regression coefficients and intercepts. Note however that the dimensions of
\(\boldsymbol{\beta}^{I}\) and \(\boldsymbol{\beta }^{II}\) need not necessarily
be the same and we are thus allowed to have a different set of predictor
variables and vectors \(\boldsymbol{z}_i^I\) and \(\boldsymbol{z}_i^{II}\) for
the two components of \(\boldsymbol{\mu}_i\). For the teacher dataset, the
regression equation for the circular outcome in the CL-PN model is
\[\hat{\boldsymbol{\mu}}_{i} = \begin{pmatrix} \mu_{i}^{I}  \\ \mu_{i}^{II}
\end{pmatrix}=\begin{pmatrix} \beta_0^{I} + \beta_1^{I}\text{SE}_i  \\
\beta_0^{II} + \beta_1^{II}\text{SE}_i \end{pmatrix}.\]

\subsubsection{The modified CL-GPN distribution}

>>>>

---
title             : "Regression models for Cylindrical data in Psychology"
shorttitle        : "Cylindrical data in Psychology"

author: 
  - name          : "Jolien Cremers"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "joliencremers@gmail.com"
  - name          : "Helena J.M. Pennings"
    affiliation   : "2,3"
  - name          : "Christophe Ley"
    affiliation   : "4"

affiliation:
  - id            : "1"
    institution   : "Department of Methodology and Statistics, Utrecht University"
  - id            : "2"
    institution   : "TNO"
  - id            : "3"
    institution   : "Department of Education, Utrecht University"
  - id            : "4"
    institution   : "Department of Applied Mathematics, Computer Science and Statistics, Ghent University"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Cylindrical data are multivariate data which consist of a directional,
  in this paper circular, and a linear component. Examples of cylindrical
  data in psychology include human navigation (direction and distance of 
  movement), eye-tracking research (direction and length of saccades) and
  data from an interpersonal circumplex (type and strength of
  interpersonal behavior). In this paper we adapt four models for
  cylindrical data to include a regression of the circular and linear
  component onto a set of covariates. Subsequently, we illustrate how to
  fit these models and interpret their results on a dataset on the
  interpersonal behavior of teachers.

keywords          : "cylindrical data, regression, interpersonal behavior"
wordcount         : "X"

bibliography      : ["CircularData.bib"]

figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf

header-includes:
  - \DeclareRobustCommand{\VANDER}[3]{#2} 
  - \usepackage{multirow}
  - \usepackage{rotating}
  - \usepackage{color}
  - \usepackage{subcaption}
   #% set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

---
```{r inlude = FALSE, echo = FALSE}

@@

```{r, echo = FALSE}
library(papaja)
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
library(caret) #cross-validation
library(coda) #geweke diagnostic
```

@@

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. @gurtman2011reasoning refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
[@leary1957;@wiggins1996history;@wubbels2006interpersonal], the circumplex of
affect [@russell1980circumplex], the circumplex of human emotion
[@plutchik1997general] or the model of human values [@schwartz1992values].
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., linear component) and direction (i.e., the linear component) are of
interest [@chrastil2017rotational] or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable)(e.g., @rayner200935th). Apart from the social
sciences data with a circular and linear outcome more commonly occur in
meteorology meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]
\newline
\indent Up until now researchers studying cylindrical data had to rely on the linear
statistical methods to analyze their research results. Focusing on only the
circular variable or only on the linear variable. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods [@gurtman2011reasoning; @pennings2017complexity;
@wright2009integrating] that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill this gap in
the literature on cylindrical models by adapting four existing cylindrical
models in such a way that they include a regression of both the linear and
circular component of a cylindrical variable onto a set of covariates. From now
on we will therefore refer to the components of the cylindrical variable as
outcome components. Secondly, we will show how using these models can benefit
the analysis of circumplex data and cylindrical data in psychology in general.
More specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the dependence
between the linear and circular component of a cylindrical variable we would
also like to predict the two components from a set of covariates in a regression
model. In our analysis, we will show how this can be done in four different ways
using the adapted cylindrical models.

\subsection{Cylindrical data}

Thus, data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g. @fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
elaborate overview).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.

@@

\section{Teacher Data}\label{Example}

@@

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
@vanderWant2015role, @Claessens2016side and @pennings2017phd. An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see @horowitz2010handbook for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. This two-dimensional space is called the interpersonal
circle/circumplex (IPC). The IPC is ``a continuous order with no beginning or
end'' [@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering
and the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable. This
circular variable represents the type of interpersonal behavior that a teacher
shows towards his/her students.\newline
\indent @Cremers2018Assessing explain the circular nature of the IPC data and
analyze them as such using a circular regression model. The two dimension
scores, Agency and Communion, can be converted to a circular score using the
two-argument arctangent function in \eqref{PredVal}\footnote{The selection of
the origin in circumplex data depends on the scaling of the Agency and Communion
scores; their respective 0 scores form the origin. Agency and Communion are
measured on a scale from 1 to 5 and for analysis purposes they are later
converted to scale ranging from -1 to 1. Scaling is however only considered an
issue in those instances where cylindrical data is derived from measurements in
bivariate space.}, where \(A\) represents a
score on the Agency dimension and \(C\) represents a score on the Communion
dimension. Note that when placing a unit circle on Figure \ref{QTI} we see that
the Agency dimension is related to the sine of the circular score and the
Communion dimension is related to the cosine of the circular score.
\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
The resulting circular variable \(\theta\) can then be modelled and takes values
in the interval $[0, 2\pi)$ or $[0^\circ, 360^\circ)$. Note that the round
brackets mean that $2\pi$ and $360^\circ$ are not included in the interval since
these represent the same value as 0 as a result of periodicity. \newline
\indent A circular analysis of circumplex data has several benefits: it is more
in line with its theoretical definition and it allows us to analyse the blend of
the two dimensions Agency and Communion instead of both dimensions separately
providing us with new insights compared to a separate analysis of the two
dimensions that is standard in the literature (see e.g. @pennings2018interpersonal, @wright2009integrating or @wubbels2006interpersonal). There is however one main drawback: when two-dimensional
data are converted to the circle we lose some information, namely the length of
the two-dimensional vector \((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of the
interpersonal behavior a teacher shows towards his/her students. In a
cylindrical model this strength (the linear outcome) can be modeled together
with the type of interpersonal behavior of a teacher (the circular outcome).
This leads to an improved analysis of interpersonal circumplex data, over either
analyzing the two dimensions separately or using a circular model, because we
take all information, circular and linear, into account. In the next section we
introduce several cylindrical models that can be used to analyze the teacher
data. First however we will provide descriptives for the teacher data.

\subsection{Data Description}\label{DataDescriptives}

```{r loaddata, return = FALSE, echo = FALSE, results = FALSE}
Dat <- read_spss("Data Heleen/mergeddata.sav")

@@

nrow(Dat)

@@

SE_low = Dat[Dat$Efficacy_CM <= 4.04,]
SE_mid = Dat[Dat$Efficacy_CM > 4.04 & Dat$Efficacy_CM < 6.04,]
SE_high = Dat[Dat$Efficacy_CM >= 6.04,]

summary(SE_low$y)
summary(SE_mid$y)
summary(SE_high$y)

plot(SE_low$y, SE_low$theta*(180/pi), 
     ylab = "IPC", xlab = "IPC strength",
     bty = "n", ylim = c(-180, 180))
points(SE_mid$y, SE_mid$theta*(180/pi), pch = 2)
points(SE_high$y, SE_high$theta*(180/pi), pch = 3)
legend(0.45, -75, legend = c("low SE", "average SE", "high SE"), pch = c(1,2,3), bty = "n")

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers. In addition to the type of interpersonal
behavior (IPC), the circular outcome, and the strength of interpersonal behavior
(IPC strength), the linear outcome, a teachers' self-efficacy (\verb|SE|)
concerning classroom management is used as covariate in the analysis.
In previous research, in psychology and education it had been shown that higher self-efficacy is related to the quality of interpersonal interactions [@locke2007selfefficacy; @want2018selfefficacy]. After listwise
deletion of missings ($3$ in total, only for the self-efficacy) we have a sample
of 148 teachers. Table \ref{Tableteacherdescriptives} shows descriptives for the
dataset. For the circular variable IPC we show sample estimates for the circular
mean $\bar{\theta}$ and concentration $\hat{\rho}$. The circular concentration
lies between 0, meaning the data is not concentrated at all \emph{i.e.} spread
over the entire circle, and 1, meaning all data is concentrated at a single
point on the circle The population values of these parameters are usually, and
also in this paper, referred to as $\mu$ (circular location) and $\kappa$
(circular concentration). For the linear variables (strength IPC and SE) we show
sample estimates of the linear mean and standard deviation. Figure
\ref{dataplot} is a scatterplot showing the relation between the linear and
circular outcome of the teacher data for teachers with low SE (below 1 sd below
the mean), average SE (between 1 sd below and 1 sd above the mean) and high SE
(above 1 sd above the mean).

@@

\begin{table}[h]
\centering
\caption{Descriptives for the teacher dataset.} 
\begin{tabular}{lrrrl}
  \noalign{\smallskip}\hline\noalign{\smallskip}
Variable & mean/$\bar{\theta}$ & sd/$\hat{\rho}$ & Range & Type \\ \hline\noalign{\smallskip}
IPC &33.22$^\circ$& 0.76 & - & Circular\\
strength IPC & 0.43 & 0.15 & 0.08 - 0.80 & Linear\\
SE & 5.04 & 1.00 & 1.5 - 7.0 & Linear\\
   \hline
\multicolumn{5}{l}{Note: For the circular variable IPC we show sample }\\
\multicolumn{5}{l}{estimates for the circular mean $\bar{\theta}$ and concentration $\hat{\rho}$.}\\
\multicolumn{5}{l}{For the linear variable we show the sample mean,}\\
\multicolumn{5}{l}{standard deviation and range.}
\end{tabular}
\label{Tableteacherdescriptives}
\end{table}

\section{Four Cylindrical Regression Models}\label{Models}

One of the goals of this paper is to show the benefits of cylindrical methods
for the analysis of circumplex data and cylindrical data in psychology in
general. To do so we decide to focus on four cylindrical models. The models were
selected for their relatively low complexity and the ease with which a
regression structure could be incorporated. But also because they show different
ways of modelling the the linear and circular outcome and thereby illustrate a
wider range of cylindrical models available in the literature. All four
cylindrical models contain a set of $q$ predictors $\boldsymbol{x} = x_1, \dots,
x_q$ and $p$ predictors $\boldsymbol{z} = z_1, \dots, z_q$ for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two models are
based on a construction by @mastrantonio2015bayesian, while the other models are
extensions of the models from @abe2017tractable and @mastrantonio2018joint. The
four cylindrical  models are introduced separately in the subsections below.
However, to provide a more succinct overview and comparison of the four models,
\ref{TableModels} gives an overview of the similarities and differences between
the models.

\begin{sidewaystable}[h]
\centering
\caption{Comparison of the four cylindrical regression models} 
\begin{tabular}{lllll}
  \noalign{\smallskip}\hline\noalign{\smallskip}
\multicolumn{1}{l}{Aspect} & CL-PN & CL-GPN  & Abe-Ley  & GPN-SSN \\ \hline\noalign{\smallskip}
$\Theta$ & &&&\\
$\:\:$Distribution& PN & GPN & Sine-skewed vM & GPN\\
$\:\:$Domain & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$\\
$\:\:$Shape & symmetric, & assymetric, & assymetric, & assymetric, \\
            & unimodal  & multimodal & unimodal   & multimodal \\\hline\noalign{\smallskip}
$Y$& &&&\\
$\:\:$Distribution & Normal & Normal & Weibull & skewed-Normal\\
$\:\:$Domain & $(-\infty, + \infty)$ & $(-\infty, + \infty)$ & $(0, + \infty)$ & $(0, + \infty)$\\
$\:\:$Shape & symmetric, & symmetric, & assymetric, & assymetric, \\
            & unimodal  & unimodal  & unimodal   & unimodal\\\hline\noalign{\smallskip}
$\Theta$-$Y$ dependence &                                   &                                   & & \\
                        & $y$ regressed on                  & $y$ regressed on                  & $\alpha$ and $\kappa$ & multivariate \\
                        & $\sin(\theta)$ and $\cos(\theta)$ & $\sin(\theta)$ and $\cos(\theta)$ & & distribution\\\hline
\multicolumn{5}{l}{Note: PN and GPN refer to the projected normal and general projected normal distribution.}\\
\multicolumn{5}{l}{vM refers to the von-Mises distribution}\\

\end{tabular}
\label{TableModels}
\end{sidewaystable}

\subsection{The Modified Circular-Linear Projected Normal (CL-PN) and Modified Circular-Linear General Projected Normal (CL-GPN) Models}\label{CL-(G)PN}

Following @mastrantonio2015bayesian we consider two models where
the relation between \(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\)
and $q$ covariates is specified as
\begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
